---
title: Simulations to assess the contribution of survival bias to protective associations of body mass index and Parkinson's disease
date: "`r Sys.Date()`"
author: Gibran Hemani
output: 
  pdf_document:
    keep_tex: true
bibliography: report.bib
---

```{r, echo=FALSE}
library(knitr)
opts_chunk$set(warning=FALSE, message=FALSE, echo=FALSE, fig.path = "images/")
read_chunk("../scripts/analysis.R")
```

```{r, echo=FALSE}
suppressPackageStartupMessages(library(dplyr))
suppressPackageStartupMessages(library(ggplot2))
suppressPackageStartupMessages(library(pander))
suppressPackageStartupMessages(library(knitr))
suppressPackageStartupMessages(library(reshape2))
suppressPackageStartupMessages(library(sn))
```

```{r setup}

```

```{r example_simulation }

```

## Summary

- Performed simulations where BMI is related to mortality to test if survival bias could induce an apparent causal association between BMI and PD
- The simulations suggest that a survival bias effect is induced, which makes it appear that higher BMI is protective of PD even when there is no real biological link between the two
- However, the estimated effects in real data are substantially larger than these induced survival bias effects, suggesting that a survival bias alone is insufficient to explain the result


## Background

Two sample Mendelian randomisation (MR) analysis has shown empirically that there is an effect of increasing body mass index (BMI) on reduced risk of Parkinson's disease (PD). One mechanism that this association could manifest without there being any underlying biological link is if individuals with high BMI have higher mortality rates, and therefore those diagnosed with PD are more likely to harbour alleles that are associated with lower BMI. 

This analysis seeks to simulate a population in which mortality is related to BMI, and PD is unrelated to BMI. MR is then performed on the sample to estimate the extent to which survival bias (or frailty) can induce an association between BMI and PD. This frailty effect is then compared against the empirical association obtained from MR.


## Simulation strategy

The basic model looks like this:

```
exposure ~ snp(s)
mortality ~ age + exposure
outcome ~ age
```

Simulate a large population ($n=500000$) where each individual has alleles at 77 BMI associated SNPs [@Locke2015]  (using only European, LD independent SNPs), PD status, age values, BMI values, and alive/dead status. Data looks like this (first 6 rows):

```{r }

kable(head(dat))

```


### Age 

The age values are generated to match the reported age distributions in [@Nalls2014]:

```{r nalls_age_dist}

```

```{r }
kable(age_summary)
```


### BMI SNPs

BMI SNPs are generated as a function of their allele frequencies, such that for individual $i$ at SNP $j$ their genotype value is $g_{ij} ~ Binom(2, p_{j})$ where $p_{j}$ is the allele frequency of SNP $j$.

### BMI values

The BMI values are a function of the BMI SNPs, such that

$$
x_{i} = \sum_{j} g_{ij} \beta_{j} + e_{j}
$$

where $e_{j} \sim N(0, \frac{V_E}{V_E + V_G})$, where the genetic variance $V_G = \sum_{j} 2p_{j}(1-p_{j})\beta_j^2$ and residual variance $V_E = V_P - V_G$. The phenotypic variance, $V_P$, is the variance of the trait that was used to obtain the effect sizes.

### PD status

PD status was simulated as a function of age, based on age related incidence obtained from Driver et al.

```{r pd_hr}

# plot of PD incidence

```

The distribution of simulated age stratified by PD looks like this:

```{r simulated_age_dist }

```


### Alive/dead status 

Alive/dead status was a function of age and BMI values. The baseline survival function was generated from the Gompertz-Makeham law of mortality, with age related hazard function

$$
h(t) = a \exp(bt) + \lambda
$$

which has CDF

$$
F(t) = 1 - \exp(-\lambda t-\frac{a}{b}(e^{b t}-1))
$$

giving the baseline survival function:

$$
S_{b}(t) = 1 - F(t) = -\exp(-\lambda t-\frac{a}{b}(e^{b t}-1))
$$

which looks like this:

```{r baseline_gm }

# plot of baseline survival

```

The influence of BMI on survival is then incorporated into the full survival model as

$$
S(t) = S_{b}(t)^{w(x)}
$$

where $x$ is the BMI value and $w(x)$ is a function that uses external data to relate BMI with mortality. A J-shaped relationship between BMI and hazard ratios [@BerringtondeGonzalez2010] was simulated:


```{r bmi_hr }

# plot of bmi - hr

```

The influence of BMI on the survival function is shown here, where the curves are the survival functions of quartiles of BMI values:


```{r plot_sim_mort_age_bmi }

# Plot of bmi survival

```

## MR analysis of simulated data

Perform an MR analysis of BMI on PD in the simulated data. A sample of `r length(index)` are randomly sampled from the individuals who have `alive` status to match the sample size and relative numbers of cases and controls in the PD GWAS [@Nalls2014]. With these simulated data the following tests can be performed:

1. Observational association between BMI and PD
2. Two stage least squares estimate of BMI on PD
3. Two sample MR using the BMI effect sizes [@Locke2015] and the estimated association between these 66 simulated SNPs and the simulated PD status. The inverse variance weighting and the MR Egger methods are shown here.

An example of what the result looks like for one simulation is shown here:

```{r example_mr }

# Example of MR analysis on the simulated data

```

Performing the simulation `r max(res$sim)` times allows us to obtain an empirical distribution of these estimates under the null model that BMI is not biologically linked with PD. The results from these simulations look like this:

```{r method_comparisons_model1 }

# Distribution of results from 1000 simulations

```

So for each of the methods (observational association, 2SLS and 2 sample MR) there is evidence that survival bias induces an apparent protective association of BMI on PD. How does this effect relate to those effects that were estimated in the real data? 

```{r empirical_results }

# Empirical comparison

```

**It is clear from this figure that the estimates obtained from the real data are substantially larger than an effect that would be induced spuriously from a frailty effect alone.**

## Limitations

The simulations rely on external data to provide hazard ratios for BMI and incidence rates for PD. The results are unlikely to be grossly effected by small fluctuations in these estimates.

The frailty simulations allow us to compare the induced associations using three different methods - observational associations, 2SLS, and 2 sample MR. The distribution of estimates from these three methods are not consistent. As an example to illustrate this more clearly, a second simulation was performed using a different hazard model for BMI. Here, a simple and extreme model was used whereby the HR for BMI on mortality was 1 for BMI values less than 27, and 5 for BMI values greater than or equal to 27. The comparison of simulation results from model 1 (J-shaped model described above) and model 2 (extreme) are shown below:

```{r method_comparisons }

```

As expected model 2 induces a much larger apparent protective effect, but it's not clear why the different methods give quite drastically different results.


## References