# Revisiting frailty

Trying to understand the actual mechanism by which frailty effects lead to incorrect MR results.

1. People with the smoking allele die quicker
2. This is a problem in observational studies, there is ascertainment in the remaining samples
3. Individuals with longer-age related phenotypes will be enriched for the non-risk allele.

This is fine. But what about when cases and controls are age matched?

Imagine two streams of individuals, those who will get the diseae (A) and those who will not (B). Within both A and B there are equal proportions of the smoking risk allele. Over time the proportion of individuals with the risk allele die quicker in each group. The only way in which you get differential genotype frequencies is if you sample from each group at different time points.

If you sample at the same time, there should be no difference. For example...

- One SNP (p=0.5)
- Case/control status is p=0.5
- Death is sampled based on SNP

```{r }
n <- 100000
dat <- data.frame(
	g = rbinom(n, 2, 0.5)
)
dat$d <- rbinom(n, 1, (dat$g+1)/4)
dat$p <- rbinom(n, 1, (dat$d+1)/3)
```

Proportions:

```{r }
with(subset(dat, d == 0), table(p))
with(subset(dat, d == 0), table(g))
with(subset(dat, d == 0), table(g, p))
```

The genotype proportions in cases and controls are identical.

Note that if this is done using a meta analysis it probably doesn't make a difference...

```{r }
nsim <- 10
n <- 10000

l <- list()
for(i in 1:nsim)
{
	dat <- data.frame(
		g = rbinom(n, 2, 0.5)
	)
	dat$d <- rbinom(n, 1, (dat$g+1)/4)
	dat$p <- rbinom(n, 1, (dat$d+1)/3)
	l[[i]] <- dat
}

dat <- plyr::rbind.fill(l)

with(subset(dat, d == 0), table(p))
with(subset(dat, d == 0), table(g))
with(subset(dat, d == 0), table(g, p))

a <- subset(dat, d == 0 & p == 0)[1:5000,]
b <- subset(dat, d == 0 & p == 1)[1:5000,]
ab <- rbind(a,b)
with(subset(ab, d == 0), table(g, p))
```

